\chapter*{Acknowledgements}

Front and foremost I would like to thank my supervisor, Thorsten
Altenkirch, for taking me on as a student and guiding me through my
PhD. I am grateful for his support and for teaching me about type
theory and doing research. He has provided me with the freedom to
follow my own ideas as well as providing necessary guidance.

Second of all, I would like to thank Natasha Alechina and Venanzio
Capretta for looking at my work and making valuable suggestions during
my first and second annual reviews respectively.

I also owe my colleagues Paolo Capriotti and Fredrik Nordvall Forsberg
a lot, as this thesis would have not existed if not for the
collaboration with them. I have learned a great deal from discussions
at the whiteboard with them.

Furthermore, I would like to thank Nicolai Kraus for his thorough
reading of this thesis, leading to fixes in proofs and readability
improvements.

The reading groups with Paolo, Nicolai Kraus, Ambrus Kaposi and Manuel
Bärenz have taught me everything I know about higher category
theory. Without these meetings I would never have reached the level of
understanding of the subject matter I have now.

I am much indebted to my examiners Bernhard Reus (external) and
Venanzio Capretta (internal) for having spent a lot of time with this
thesis. They have provided with valuable feedback that has improved
the thesis considerably.

The other members of the Functional Programming Lab have also greatly
contributed to making my PhD experience an enjoyable one: Graham,
Henrik, Christian, Florent, Laurence, Bas, Nuo, Jenny, Iv\'an, Jan,
Jon and Jakob.

Everybody who visited me in Nottingham or received me with open arms
when I visited them, deserves a very big thanks: Paul, Justin, Joshua,
Robert, Wout, Edo, Ingo, Felix, Wout, Jonas, Koen, Marnix and
countless of people I forgot to mention.

Last but not least, I would like to thank my parents and Rehma for
their unconditional support.
